---
title: "Durability and recovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Durability and recovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, cache = FALSE, collapse = TRUE,
                    comment = "#>", error = FALSE, message = FALSE)
```

## What survives and what does not

[Getting started](getting-started.html#the-concepts) defines a task, handler,
attempt, step, checkpoint, and lease. This guide describes what those concepts
mean when a worker, process, or connection fails.

CanardAbsurd provides at-least-once task execution with durable named
checkpoints. A task input, state, lease, counters, and completed step values
survive a worker failure. A replacement worker resumes the handler and replays
those completed steps. Completed, failed, and cancelled tasks are terminal.

This is not exactly-once execution of external effects. A worker can call an
external service and fail before its step result is saved; a later attempt then
calls that service again. Use an application-level idempotency key for every
external mutation.

For a shared workflow database, run one Quack server process as the writable
DuckDB owner. Producers and workers connect through Quack rather than opening
the database file themselves, avoiding direct multi-process DuckDB/WAL locking
and errors.

## Run the examples

The examples use `ca_open()` only to inspect the state machine without a network
or Quack installation. This in-memory database belongs to this one R process;
it is not a shared deployment.

```{r open}
library(CanardAbsurd)
db <- ca_open()
withr::defer(ca_close(db))
```

## How task state changes

```mermaid
stateDiagram-v2
  direction LR
  [*] --> ready: submit
  ready --> running: claim
  running --> completed: complete
  running --> ready: retry / sleep
  running --> running: recover
  running --> failed: exhausted
  ready --> cancelled: cancel
  running --> cancelled: cancel
```

Each state transition is one server-side SQL statement; the R handler never
runs inside its transaction. A network failure does not prove that a request was
never executed, so it is surfaced rather than retried. A known write conflict is
different: DuckDB aborted the statement, so repeating it is safe. Competing
workers conflict routinely because they select the same eligible row.
`ca_work()` therefore repeats each conflicting statement up to
`conflict_retries` times (8 by default) with jittered exponential backoff,
without rerunning an R callback. Outside a worker, callers handle
`canard_retryable` and invoke its `canard_retry` restart themselves; without
that decision the conflict propagates.

Recovery replaces an expired lease when the failure budget allows another
attempt. The `attempt` counter counts claims, including sleep resumptions. The
`failures` counter counts reported failures and expired leases. `max_failures`
is the threshold for terminal failure. Before claiming, `ca_claim()` and
`ca_work()` mark a bounded batch of expired, exhausted tasks as failed in their
selected queue. `reap_limit` defaults to 64. An expired task with failure budget
remaining can instead be reclaimed by a worker; expiry does not by itself start
another attempt.

## Native lease maintenance

A queue with no polling workers still needs cleanup when its final attempt
expires. The optional `canard_coordinator` extension performs that cleanup across
all queues, on a dedicated connection and native thread in the database-owner
process. It increments the failure count, records `worker lease expired`, clears the
worker/token/deadline, and marks an exhausted task failed in one SQL statement.

```mermaid
flowchart LR
  Poll["worker polls its queue"] --> Reap["mark expired final attempts failed"]
  Native["optional native maintenance<br/>polls all queues"] --> Reap
  Reap --> DB[("DuckDB task rows")]
  Poll --> Claim["claim eligible work<br/>or reclaim an expired retryable task"]
  Claim --> Handler["R worker runs the handler"]
```

Use `ca_coordinator_start()` on the `ca_open()` or `ca_serve()` owner handle, not
a remote `ca_connect()` client. `poll_milliseconds` sets the maintenance interval;
`reap_limit` bounds the tasks changed by each poll, not the amount of table data
DuckDB may scan. The native loop does not replace worker polling and does not
renew a worker's lease, run a handler, enforce CPU/memory admission, or decide
which downstream tasks are ready.

`ca_coordinator_status()` reports state, poll/reaped/error counters, and the last
error. Starting the loop is not proof that maintenance SQL succeeds; inspect
those counters and errors. `ca_coordinator_stop()` joins the thread and releases
its connection. `ca_close()` does this whenever the handle loaded the extension,
and still does it if stopping the Quack endpoint fails. The instance is
one-shot: after stop, use a fresh database instance to start another
coordinator. The extension holds its connection from the moment it loads, so
closing the DBI connection directly, bypassing `ca_close()`, keeps the database
open and locked until the process exits.

The extension is built separately against the stable DuckDB C extension API.
Loading and starting it are explicit; opening an ordinary database does neither.
No signed build is distributed. A local build loads only into a database opened
with `allow_unsigned_extensions = TRUE`, a development-only setting that lets
any SQL on that database load native code. The native path is experimental. See
the [extension instructions](https://github.com/RGenomicsETL/CanardAbsurd/tree/main/tools/canard-coordinator)
for build and executable native integration tests.

## A replaced claim cannot write

Failure releases the original claim. A subsequent claim receives a new token;
completion with the old token is rejected without changing the current task.

These examples share a local database opened with `ca_open()`.

```{r fencing}
id <- ca_spawn(db, "example", id = "lease-example")
old <- ca_claim(db)
ca_fail(old, "retry")
current <- ca_claim(db)

outcome <- tryCatch({
  ca_complete(old, "stale")
  "accepted"
}, canard_lease_lost = function(e) "rejected")
ca_complete(current, "current")
task <- ca_inspect(db, id)
list(stale_completion = outcome, result = task$result)
```

Heartbeats, checkpoints, completion, failure, and suspension all require the
current unexpired token. An expired lease cannot be revived by heartbeating.
Cancellation also invalidates subsequent worker writes; it does not interrupt
an external operation already in progress.

## Checkpoints preserve native values

Vectors, lists, data frames, dates, timestamps and other supported R values
retain their declared mappings on execution and replay. A saved `NULL` is
distinct from a missing checkpoint. See [Native values and DuckDB R compatibility](native-values.html)
for the supported classes, precision limits and driver compatibility details.

```{r null-checkpoint}
id <- ca_spawn(db, "example")
first <- ca_claim(db)
ca_step(first, "nullable", function() NULL)
ca_fail(first, "retry")
second <- ca_claim(db)
value <- ca_step(second, "nullable", function() stop("must replay"))
ca_complete(second, value)
list(replayed_null = is.null(value), state = ca_inspect(db, id)$state)
```

Step names must be unique within an attempt. Include a stable index for loop
steps. Treat step names and result shapes as persistent interfaces when deploying
new handler code.

## Query workflow state with SQL

Task state, queue, attempts, failures, and timestamps are typed columns. Inputs
and results are native VARIANT values with separate R type descriptors.
Checkpoint names, kinds and native values are available through
`canard_absurd.task_checkpoints`.

```{r sql-inspection}
DBI::dbGetQuery(db@con, "
  SELECT state, count(*) AS tasks
  FROM canard_absurd.tasks
  GROUP BY state
  ORDER BY state
")
DBI::dbGetQuery(db@con, "
  SELECT name, kind, variant_typeof(value) AS value_type
  FROM canard_absurd.task_checkpoints
  WHERE task_id = ?
", params = list(id))
```

These queries need no DuckDB extensions. Nested fields can be extracted and cast
with native SQL. On affected engines, materialize whole values before indexed
field extraction; the compatibility guide includes the query form and an
executable explanation of the separate nested-NULL conversion issue.

## Inspect an attempt's failure

A handler failure and a task's terminal failure are different events. `ca_run()`
returns both the attempt status and the persisted task state, along with the
original R condition. Persistence errors and interrupts propagate instead of
being recorded as handler failures.

```{r handler-failure}
id <- ca_spawn(db, "example", max_failures = 1)
outcome <- ca_run(ca_claim(db), function(input, ctx) stop("delivery unavailable"))
list(status = outcome$status, state = outcome$state,
     error = conditionMessage(outcome$error))
```

If saving a handler failure also fails, `canard_failure_recording_error` retains
the handler condition in `parent` and the storage condition in
`persistence_error`. Error text is stored without truncation. Rich R conditions
remain in attempt outcomes; the database stores their messages, not serialized R
objects. See `?ca_conditions` for the condition fields and retry notifications.

## Error metadata compatibility

Error preservation depends on the installed R driver and Quack build. Affected
R wrappers discard the original condition's fields and cause chain; affected
Quack bind paths turn the transported structured error into an
`InvalidInputException`.

The R wrapper report is [duckdb-r #2711](https://github.com/duckdb/duckdb-r/issues/2711).
The fixes in [duckdb-r #2714](https://github.com/duckdb/duckdb-r/pull/2714) and
[duckdb-quack #212](https://github.com/duckdb/duckdb-quack/pull/212) are merged.
Their merge status alone does not establish that an installed driver/extension
pair includes both fixes.

CanardAbsurd first uses a structured DuckDB transaction error when available.
Otherwise, one compatibility adapter recognizes exact tested write-conflict
messages and duplicate-key messages containing the submitted task ID. It does
not search arbitrary diagnostic text for conflict fragments. A
`canard_conflict` or `canard_retryable` condition has `message_based = TRUE` when
that adapter supplied the classification, and `FALSE` when structured metadata
supplied it. Structured non-transaction types take precedence over the text.

This is a workaround for released drivers, **not a stable error protocol**.
Changed or unknown message forms propagate without a retry restart, as do
transport failures. An unhandled recognized conflict also propagates: retry
limits and delays still belong to the caller. Revalidate the DuckDB/Quack pair
when upgrading; the integration suite uses installed packages and independent
R processes. The package requires DuckDB R 1.5.5 or newer. Historical probes on
1.5.3 explain the compatibility adapter, but that version is not a supported
package runtime.

## External effects still need idempotency

A worker can call an external service, crash before saving the result, and repeat
that call on recovery. Checkpointing does not make an HTTP request transactional
with DuckDB. Use a stable business-level idempotency key for payments, messages,
and other mutations. Deduplicating task submission does not deduplicate those
external effects.

## Supervise work longer than a lease

Calls to `ca_step()`, `ca_sleep()` and `ca_heartbeat()` renew a lease, but no
background R thread renews it while a callback blocks. A step that blocks in
`system2()` longer than its lease lets another worker reclaim the task and start
the same work again. The database still rejects the old worker's writes, but not
its files.

`ca_process()` runs the command in its own process and renews the lease from
the worker while it runs. It checks the lease before starting, writes logs to
`stdout.log` and `stderr.log`, and gives each attempt its own `attempt-<n>`
directory, so an obsolete attempt cannot overwrite accepted files. Lease loss,
including cancellation, a timeout, or an error in the supervising R code kills
the command's process tree. Returning its paths from a `ca_step()` callback
publishes them only while the lease is held. For blocking R or native
computation, run an `Rscript` the same way.

```{r process}
rscript <- file.path(R.home("bin"),
  if (.Platform$OS.type == "windows") "Rscript.exe" else "Rscript")
root <- tempfile("canard-work-")
withr::defer(unlink(root, recursive = TRUE))
align <- function(input, ctx) {
  ca_step(ctx, "align", function() {
    directory <- ca_process(ctx, rscript,
      c("-e", "Sys.sleep(1.5); writeLines('aligned', 'sample.bam')"),
      directory = file.path(root, ctx@id, "align"), heartbeat_seconds = 0.2)
    file.path(directory, "sample.bam")
  })
}
id <- ca_spawn(db, "align", id = "sample-101")
ca_work(db, list(align = align), max_tasks = 1, lease_seconds = 0.5)
task <- ca_tasks(db, id = id)
list(state = task$state, attempts = task$attempt,
     output = basename(dirname(ca_result(db, id))))
```

The command ran three times longer than the half-second lease without a
reclaim. Supervision cannot rule out every physical overlap: a paused or
partitioned worker can outlive its lease before its next heartbeat. For Slurm
or another scheduler, use that backend's own submission, discovery and
cancellation identities; supervising a submit command does not supervise the
remote job.

## Operational limits

- Checkpoints share the task row so lease changes and checkpoint writes contend
  on the same durable record. This keeps lease ownership and checkpoint storage
  in one update; it is not a throughput guarantee.
- Inputs, results, checkpoints and error text have no package-imposed byte quota.
  Large values still consume engine memory, disk and transport capacity.
  Values travel as SQL literals, and each new checkpoint rewrites the task's
  checkpoint map, so cost grows with the number and size of saved steps. Store
  file references rather than bulk data, and measure a workload with
  [`tools/benchmark.R`](https://github.com/RGenomicsETL/CanardAbsurd/blob/main/tools/benchmark.R)
  before relying on many large checkpoints.
- `ca_tasks()` lists metadata without reading payloads; use it to poll many
  tasks. `ca_inspect()` reads every value of one task.
- Claims and checkpoint replay use a fenced typed read after their lease UPDATE.
  A failure of that read does not undo the UPDATE or cause a handler to run.
  See the compatibility guide for the additional round trip and recovery rules.
- Schema version 1 bounds task IDs and handler names to 256 characters, queue
  names to 128 characters, and failure budgets to 1 through 1,000,000. Changing
  those persisted constraints requires a schema change, not a client option.
- Task IDs remain deduplicated while their rows are retained. Retention and
  database growth are operator responsibilities.
- Schema version 1 is checked at open/connect time. Events, cron, automatic
  retention, history tables, and schema migrations are outside this release.
- Quack is experimental. Record `ca_runtime()` when validating a deployment;
  compatibility depends on the DuckDB and Quack versions in deployment.

## Close the example database

```{r close}
ca_close(db)
```
