---
title: "Durability and recovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Durability and recovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, cache = FALSE, collapse = TRUE,
                    comment = "#>", error = FALSE, message = FALSE)
library(CanardAbsurd)
db <- ca_open()
withr::defer(ca_close(db))
```

## What is durable?

DuckDB stores each task's inputs, state, lease token, counters, and checkpoint
map. The R handler runs outside a database transaction. Each state transition
is a server-side SQL statement. Callers can handle `canard_retryable` and invoke
its `canard_retry` restart to repeat a known aborted write conflict. Without that
explicit decision, the error propagates. A network failure does not prove that
a request was never executed.

```mermaid
stateDiagram-v2
  direction LR
  [*] --> ready: submit
  ready --> running: claim
  running --> completed: complete
  running --> ready: retry / sleep
  running --> running: recover
  running --> failed: exhausted
  ready --> cancelled: cancel
  running --> cancelled: cancel
```

Completed, failed, and cancelled tasks are terminal. Recovery replaces an expired
lease when the failure budget allows another attempt.

The `attempt` counter counts claims, including sleep resumptions. The `failures`
counter counts reported failures and expired leases. `max_failures` is the
threshold for terminal failure. Reaping happens on the claim path, so an expired
last attempt is marked failed when workers next poll its queue. `reap_limit`
controls the size of that maintenance batch; its default is 64 rows.

## A replaced claim cannot write

Failure releases the original claim. A subsequent claim receives a new token;
completion with the old token is rejected without changing the current task.

These examples share a local database opened with `ca_open()`.

```{r fencing}
id <- ca_spawn(db, "example", id = "lease-example")
old <- ca_claim(db)
ca_fail(old, "retry")
current <- ca_claim(db)

outcome <- tryCatch({
  ca_complete(old, "stale")
  "accepted"
}, canard_lease_lost = function(e) "rejected")
ca_complete(current, "current")
task <- ca_inspect(db, id)
list(stale_completion = outcome, result = task$result)
```

Heartbeats, checkpoints, completion, failure, and suspension all require the
current unexpired token. An expired lease cannot be revived by heartbeating.
Cancellation also invalidates subsequent worker writes; it does not interrupt
an external operation already in progress.

## Checkpoints are JSON values

JSON arrays normalize to R lists on both execution and replay. JSON objects
become named lists. A saved `NULL` is distinct from a missing checkpoint.

```{r null-checkpoint}
id <- ca_spawn(db, "example")
first <- ca_claim(db)
ca_step(first, "nullable", function() NULL)
ca_fail(first, "retry")
second <- ca_claim(db)
value <- ca_step(second, "nullable", function() stop("must replay"))
ca_complete(second, value)
list(replayed_null = is.null(value), state = ca_inspect(db, id)$state)
```

Step names must be unique within an attempt. Include a stable index for loop
steps. Treat step names and result shapes as persistent interfaces when deploying
new handler code.

## Inspect an attempt's failure

A handler failure and a task's terminal failure are different events. `ca_run()`
returns both the attempt status and the persisted task state, along with the
original R condition. Persistence errors and interrupts propagate instead of
being recorded as handler failures.

```{r handler-failure}
id <- ca_spawn(db, "example", max_failures = 1)
outcome <- ca_run(ca_claim(db), function(input, ctx) stop("delivery unavailable"))
list(status = outcome$status, state = outcome$state,
     error = conditionMessage(outcome$error))
```

If saving a handler failure also fails, `canard_failure_recording_error` retains
the handler condition in `parent` and the storage condition in
`persistence_error`. Error text is stored without truncation. Rich R conditions
remain in attempt outcomes; the database stores their messages, not serialized R
objects. See `?ca_conditions` for the condition fields and retry notifications.

## Error metadata compatibility

Two upstream boundaries lose structured information:

- [DuckDB R #2711](https://github.com/duckdb/duckdb-r/issues/2711): the DBI rethrow
  discards the original error fields and parent.
- [Quack PR #212](https://github.com/duckdb/duckdb-quack/pull/212): the proposed
  protocol change preserves the server's error type and transferable metadata.

CanardAbsurd first uses a structured DuckDB transaction error when available.
Otherwise, one compatibility adapter recognizes exact tested write-conflict
messages and duplicate-key messages containing the submitted task ID. It does
not search arbitrary diagnostic text for conflict fragments. A
`canard_conflict` or `canard_retryable` condition has `message_based = TRUE` when
that adapter supplied the classification, and `FALSE` when structured metadata
supplied it. Structured non-transaction types take precedence over the text.

This is a workaround for released drivers, **not a stable error protocol**.
Changed or unknown message forms propagate without a retry restart, as do
transport failures. An unhandled recognized conflict also propagates: retry
limits and delays still belong to the caller. Revalidate against the actual
DuckDB/Quack pair when upgrading; the integration suite uses installed packages
and independent R processes. Tests exercise DuckDB R 1.5.3 and 1.5.5 and the
community Quack builds available for them.

## External effects still need idempotency

A worker can call an external service, crash before saving the result, and repeat
that call on recovery. Checkpointing does not make an HTTP request transactional
with DuckDB. Use a stable business-level idempotency key for payments, messages,
and other mutations. Deduplicating task submission does not deduplicate those
external effects.

Long operations must call `ca_heartbeat(ctx)` before their lease expires. Step
boundaries renew leases, but no background R thread renews a lease while a
callback blocks.

## Operational boundaries

- Checkpoints share the task row so lease changes and checkpoint writes contend
  on the same durable record. This trades whole-row JSON updates for a simple
  ownership invariant, not a throughput guarantee.
- Schema version 1 limits input to 1 MiB of UTF-8 JSON and the accumulated
  checkpoint document to 16 MiB, including escaped value JSON and record keys.
  Results and error text have no additional package-imposed size cap.
- Schema version 1 bounds task IDs and handler names to 256 characters, queue
  names to 128 characters, and failure budgets to 1 through 1,000,000. Changing
  those persisted constraints requires a schema change, not a client option.
- Task IDs remain deduplicated while their rows are retained. Retention and
  database growth are operator responsibilities.
- Schema version 1 is checked at open/connect time. Events, cron, automatic
  retention, history tables, and schema migrations are outside this release.
- Quack is experimental. Record `ca_runtime()` when validating a deployment;
  compatibility depends on the actual DuckDB and Quack pair.
