---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, cache = FALSE, collapse = TRUE, comment = "#>", error = FALSE)
library(CanardAbsurd)
```

A workflow is an R function taking `(input, ctx)`. `ca_spawn()` stores its name
and JSON input; `ca_work()` finds the corresponding function in a named list.

This guide uses an in-memory database. All examples execute during rendering,
without network access or a Quack installation. The same workflow functions run
against a remote connection made with `ca_connect()`.

## Run a task

```{r first-task}
local({
  db <- ca_open()
  on.exit(ca_close(db))

  id <- ca_spawn(db, "total", list(amount = 20, tax = 2), id = "order-001")
  handlers <- list(total = function(input, ctx) {
    ca_step(ctx, "calculate", function() input$amount + input$tax)
  })

  ca_work(db, handlers, max_tasks = 1)
  task <- ca_inspect(db, id)
  stopifnot(task$state == "completed", task$result == 22)
  list(state = task$state, result = task$result)
})
```

The stable ID also makes submission idempotent: submitting the same name, input,
queue, priority, and failure budget under that ID returns the existing task.
Different input under that ID raises `canard_spawn_conflict`. JSON object key
order is part of submission equality.

## Replay after a failure

The first step below completes once. The second step fails on its first call,
so the worker retries the handler. The saved first step is replayed.

```{r replay}
local({
  db <- ca_open()
  on.exit(ca_close(db))
  calls <- c(prepare = 0L, deliver = 0L)

  handler <- function(input, ctx) {
    value <- ca_step(ctx, "prepare", function() {
      calls[["prepare"]] <<- calls[["prepare"]] + 1L
      input$value * 2
    })
    ca_step(ctx, "deliver", function() {
      calls[["deliver"]] <<- calls[["deliver"]] + 1L
      if (calls[["deliver"]] == 1L) stop("temporary failure")
      value
    })
  }

  id <- ca_spawn(db, "replay", list(value = 21), id = "replay-001")
  ca_work(db, list(replay = handler), idle_timeout = 0)
  task <- ca_inspect(db, id)
  stopifnot(task$state == "completed", task$result == 42,
            identical(calls, c(prepare = 1L, deliver = 2L)))
  data.frame(state = task$state, attempts = task$attempt,
             failures = task$failures, prepare_calls = calls[["prepare"]],
             deliver_calls = calls[["deliver"]], result = task$result)
})
```

The local counters expose which callbacks ran; they are not durable workflow
state. Actual durable results live in the database.

## Sleep without occupying a worker

`ca_sleep()` saves its checkpoint and releases the claim. The resumed handler
passes that named sleep without scheduling it again.

```{r sleep}
local({
  db <- ca_open()
  on.exit(ca_close(db))
  id <- ca_spawn(db, "pause", max_failures = 1)
  handler <- function(input, ctx) {
    ca_sleep(ctx, "pause", seconds = 0)
    "resumed"
  }

  ca_work(db, list(pause = handler), max_tasks = 2)
  task <- ca_inspect(db, id)
  stopifnot(task$result == "resumed", task$attempt == 2L, task$failures == 0L)
  list(result = task$result, attempts = task$attempt, failures = task$failures)
})
```

A zero-second sleep demonstrates the state transition without delaying the
render. Positive durations use the database clock. Sleeps do not consume the
failure budget.

## Choose a worker lifetime

- `max_tasks = 1` executes one claimed attempt, including a failure or suspension.
- `idle_timeout = 0` drains work that is currently eligible.
- The defaults keep polling until the worker is interrupted.

Each worker runs one attempt at a time. For concurrency, run independent R
processes connected to a Quack server. Only names present in `handlers` are
claimed. See the [Quack server guide](https://rgenomicsetl.github.io/CanardAbsurd/articles/quack-server.html).
