---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, cache = FALSE, collapse = TRUE,
                    comment = "#>", error = FALSE, message = FALSE)
```

A task is a durable request to run a named R handler. The handler itself stays
in worker code; the database stores the request, its state, and values saved by
its named steps.

## The concepts

- **Task:** one persistent request. It has a stable ID, handler name, input,
  queue, state, attempts, and failure budget.
- **Handler:** an R function taking `(input, task)`. A worker supplies it from a
  named handler list; it is not stored in the database.
- **Attempt:** one worker's claim and execution of a handler. A replacement
  worker starts another attempt after failure, expiry, or a durable sleep.
- **Step:** a named point in a handler where `ca_step()` saves a value. The name
  is a persistent part of the handler's contract.
- **Checkpoint:** the value saved for a completed step. A later attempt receives
  it from `ca_step()` without running that step callback again.
- **Lease:** the worker's temporary right to update the claimed task. A stale or
  cancelled worker cannot checkpoint or complete it.

`ca_spawn()` creates a task. `ca_work()` claims tasks for the handler names it
knows, executes attempts, and reports their outcomes. `ca_inspect()` reads the
current task record.

This guide uses an in-memory database to introduce that lifecycle without
network access or a Quack installation. It is a one-process example: no other
R process accesses this database. For a shared workflow database, Quack is
mandatory: one server process owns the writable DuckDB file, and producers and
workers use `ca_connect()` rather than opening that file.

## Submit, work, and inspect one task

```{r open}
library(CanardAbsurd)
db <- ca_open()
withr::defer(ca_close(db))
```

```{r first-task}
total <- function(input, ctx) {
  ca_step(ctx, "calculate", function() input$amount + input$tax)
}

id <- ca_spawn(db, "total", list(amount = 20, tax = 2), id = "order-001")
ca_work(db, list(total = total), max_tasks = 1)
task <- ca_inspect(db, id)
list(state = task$state, result = task$result)
```

The stable ID also makes submission idempotent: submitting the same name, input,
queue, priority, and failure budget under that ID returns the existing task.
Different input under that ID raises `canard_spawn_conflict`. Input equality
includes the R mapping, names, and their order.

## Resume after a failure

The first step below completes once. The second step fails on its first call,
so the worker retries the handler. The saved first step is replayed.

```{r replay}
calls <- c(prepare = 0L, deliver = 0L)
replay <- function(input, ctx) {
  value <- ca_step(ctx, "prepare", function() {
    calls[["prepare"]] <<- calls[["prepare"]] + 1L
    input$value * 2
  })
  ca_step(ctx, "deliver", function() {
    calls[["deliver"]] <<- calls[["deliver"]] + 1L
    if (calls[["deliver"]] == 1L) stop("temporary failure")
    value
  })
}

report <- function(outcome) {
  if (!is.null(outcome$error)) {
    cat("Attempt", outcome$attempt, ":", conditionMessage(outcome$error), "\n")
  }
}
id <- ca_spawn(db, "replay", list(value = 21), id = "replay-001")
ca_work(db, list(replay = replay), idle_timeout = 0, on_result = report)
task <- ca_inspect(db, id)
data.frame(state = task$state, attempts = task$attempt,
           failures = task$failures, prepare_calls = calls[["prepare"]],
           deliver_calls = calls[["deliver"]], result = task$result)
```

`on_result` receives every persisted attempt outcome, including the original
handler condition on failure. Without a callback, the worker warns on failures.
`failure_delay` controls when a failed task becomes eligible again; its default
is zero. The local counters expose which callbacks ran; they are not durable
workflow state. Durable results live in the database.

## Release a task until later

`ca_sleep()` saves its checkpoint and releases the claim. The resumed handler
passes that named sleep without scheduling it again.

```{r sleep}
pause <- function(input, ctx) {
  ca_sleep(ctx, "pause", seconds = 0)
  "resumed"
}
id <- ca_spawn(db, "pause", max_failures = 1)
ca_work(db, list(pause = pause), max_tasks = 2)
task <- ca_inspect(db, id)
list(result = task$result, attempts = task$attempt, failures = task$failures)
```

A zero-second sleep demonstrates the state transition without delaying the
render. Positive durations use the database clock. Sleeps do not consume the
failure budget.

## Choose a worker lifetime

- `max_tasks = 1` executes one claimed attempt, including a failure or suspension.
- `idle_timeout = 0` drains work that is currently eligible.
- The defaults keep polling until the worker is interrupted.

Each worker runs one attempt at a time. For concurrency, run independent R
processes connected to a Quack server. Only names present in `handlers` are
claimed. See the [Quack server guide](https://rgenomicsetl.github.io/CanardAbsurd/articles/quack-server.html).

Close the one-process connection when finished. The deferred cleanup also
closes it if a later example errors.

```{r close}
ca_close(db)
```
