---
title: "Native values and DuckDB R compatibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Native values and DuckDB R compatibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Storage and R mappings

CanardAbsurd stores input, result, and checkpoint values as native DuckDB
`VARIANT` values. It does not impose input or cumulative checkpoint byte quotas.
DuckDB memory, disk, and transport limits still apply; keeping a large value in
the task row does not make copying or transferring it free.

R type descriptors accompany the values. They distinguish `NULL` from typed
missing values, vectors from lists, and data frames from named lists. They also
retain underlying R vector storage, names, factor levels, time zones and row names. See `?ca_values` for the
supported classes and their mappings. Timestamps and durations have DuckDB
microsecond precision. Custom classes require explicit conversion.

The package requires DuckDB R 1.5.5 or newer. Earlier engines also have a
reported failure when appending to indexed tables after VARIANT shredding
([DuckDB #25024](https://github.com/duckdb/duckdb/issues/25024)). The file format
must support persisted VARIANT values. `ca_open()` requests DuckDB storage
compatibility v1.5.0. This is separate from CanardAbsurd's own
schema version, which remains 1. These unreleased schema-1 files require matching
package code; no schema migration is provided.

## Field-name admission

Named lists and data frame columns require nonempty names that are unique
ignoring ASCII letter case. They are encoded through DuckDB STRUCT fields, so
`A` and `a` cannot coexist in one container. Names are stored unchanged;
non-ASCII letters are not case-folded. Atomic vector names are not subject to
this restriction.

These unsupported containers raise `canard_value_error` before submission SQL
or checkpoint persistence. If a callback returns one, `ca_run()` records a
handler failure rather than treating it as a storage failure. Callers needing
case-colliding field names can explicitly serialize the object to a BLOB; see
[Serialized R values](serialized-values.html).

```{r field-name-admission}
library(CanardAbsurd)
db <- ca_open()
tryCatch(
  ca_spawn(db, "example", list(A = 1L, a = 2L)),
  canard_value_error = conditionMessage
)
ca_close(db)
```

## Nested NULL conversion in released R drivers

DuckDB R 1.5.3 and 1.5.5 fail to convert some nested VARIANT values to R. This is
reproducible with DBI alone: it does not require CanardAbsurd, Quack, JSON, or a
file-backed database. An all-NULL VARIANT array and an object with a NULL field
produce diagnostics such as:

```
Unknown type for column `variant`: "NULL"
Context: LogicalTypeId::LIST

Unknown type for column `variant$x`: "NULL"
Context: LogicalTypeId::STRUCT
```

The SQL values are valid. Reading a scalar VARIANT NULL works, as do ordinary
typed SQL arrays and structs containing missing values. It is the nested
VARIANT-to-R conversion that fails.

This reproducible example reports the behavior of the installed driver. Errors
are caught so both failing cases and their typed controls are evaluated.

```{r connection}
library(DBI)
con <- dbConnect(duckdb::duckdb(),
  config = list(autoinstall_known_extensions = "false"))
packageVersion("duckdb")
dbGetQuery(con, "PRAGMA version")
```

```{r nested-null-reproducer}
queries <- c(
  scalar_null = "SELECT NULL::VARIANT AS value",
  variant_array = "SELECT [NULL, NULL]::VARIANT AS value",
  typed_array = "SELECT ([NULL, NULL]::VARIANT)::INTEGER[] AS value",
  variant_object = "SELECT {'x': NULL}::VARIANT AS value",
  typed_object = "SELECT ({'x': NULL}::VARIANT)::STRUCT(x INTEGER) AS value"
)
lapply(queries, function(sql) {
  tryCatch(dbGetQuery(con, sql), error = function(error) conditionMessage(error))
})
```

```{r close-connection}
dbDisconnect(con, shutdown = TRUE)
```

The following configurations were also checked in isolated R processes:

| R driver source/runtime | Nested VARIANT NULLs | Typed projections |
|:--|:--|:--|
| DuckDB R 1.5.3 | Conversion error | Pass |
| DuckDB R 1.5.5 | Conversion error | Pass |
| R fix `383dc751`, engine 1.5.5 | Conversion error | Pass |
| R branch commit `013d4b59`, engine `c3c91aa84f` (2.0-dev) | Pass | Pass |

The 2.0 development R build was installed under the canonical package name for
isolated testing; its package-version label is `1.5.5.9020.1456`, not its engine
version. That result is not evidence that a released 1.5 driver contains the fix.

In the affected source, the VARIANT conversion calls
`VariantUtils::ConvertVariantToValue()` followed by `RApiTypes::ValueToSexp()`.
The R allocation switch lacks a `LogicalTypeId::SQLNULL` case. Compare
[`transform.cpp` at 383dc751](https://github.com/duckdb/duckdb-r/blob/383dc751dd138685a1d305063b570188a339dd1a/src/transform.cpp)
with [the tested 2.0 branch source](https://github.com/duckdb/duckdb-r/blob/013d4b59cab52cc23927a390de2a14952ef7ac19/src/transform.cpp),
which handles SQLNULL during allocation and conversion. That handling was added
in [R branch commit feeca1e1](https://github.com/duckdb/duckdb-r/commit/feeca1e1ae289cbffec4b93f8ba0d08bc9f3e422),
which vendors DuckDB's general SQLNULL result support. The structured-error fix
in DuckDB R #2714 addresses a different problem.

## Indexed VARIANT field extraction

There is a separate engine issue, [DuckDB #24064](https://github.com/duckdb/duckdb/issues/24064):
pushing VARIANT field extraction into an index scan can cause an internal error
and invalidate the database. The upstream report also describes incorrect NULL
results after a file checkpoint and reopen. The error is in
`VariantColumnData::FetchRow()`, not R conversion or Quack transport. It is
reproducible on the tested 1.5.3 and 1.5.5 engines; the tested 2.0 development
runtime completes the same CanardAbsurd inspection sequence.

CanardAbsurd reads whole values into a `MATERIALIZED` CTE before applying typed
field projections. This retains the indexed task-ID lookup while keeping field
extraction outside the table's index scan. It does not disable indexes or change
global optimizer settings. Materialization has a memory/copying cost, which is
in addition to transferring the result to R.

Direct SQL consumers on affected engines need the same query shape. A type cast
alone does not prevent the index-scan issue:

```{r indexed-value-query}
con <- dbConnect(duckdb::duckdb())
dbExecute(con, "CREATE TABLE events (id INTEGER PRIMARY KEY, value VARIANT)")
dbExecute(con, "INSERT INTO events VALUES (1, {'duration_ms': 100}::VARIANT)")
dbGetQuery(con, "
  WITH payload AS MATERIALIZED (
    SELECT value FROM events WHERE id = 1
  )
  SELECT value['duration_ms']::DOUBLE AS duration_ms FROM payload
")
dbDisconnect(con, shutdown = TRUE)
```

A closed upstream issue or a passing development runtime does not establish that
a released engine contains the fix. Regression coverage must include indexed
reads before and after checkpoint/reopen, as well as appends after shredding.

## Why claims and replay perform a typed read

CanardAbsurd uses the recorded R descriptor to construct a SQL projection with
concrete result types. For example, an all-missing integer vector is fetched as
`INTEGER[]`, not handed to the driver's generic VARIANT decoder. The descriptor
also preserves R distinctions that cannot be inferred from a NULL value alone.

The operation order and consequences are explicit:

1. **Claim:** one UPDATE acquires the lease and returns the input descriptor.
   A second, read-only statement fetches the typed input, requiring that same
   token and an unexpired lease. No handler runs before that read succeeds.
2. **Checkpoint replay:** the entry UPDATE checks ownership, renews the lease
   and returns the requested checkpoint's descriptor. A fenced SELECT reads the
   saved value. The callback is not rerun to recover from a read failure.
3. **New checkpoint and completion:** the descriptor is already known from the
   supplied R value. The single UPDATE stores it with the value. Checkpoints and
   `ca_run()` return a typed projection; `ca_complete()` returns only an
   acknowledgement without the payload. No R callback runs inside that
   statement's transaction.

The extra SELECT adds a round trip, including over Quack. It is not part of the
preceding UPDATE's transaction. A read failure does not undo a successful claim
or lease renewal. Cancellation or expiry between the statements produces
`canard_lease_lost`; a transport failure is surfaced. A worker that disappears
after claiming leaves a lease for the normal expiry/reclaim mechanism.

Caller-controlled SQL retries remain statement-scoped: retrying the typed SELECT
does not repeat the claim UPDATE or an R handler. Unknown or ambiguous transport
failures do not acquire an automatic retry policy.

Inspection first reads state and checkpoint membership, then retrieves the
referenced values. Input, saved checkpoints and completed results are immutable
under the package protocol, so this retains the first query's view even if a
worker appends a checkpoint or completes between the reads. Direct edits to
stored payloads violate that assumption.

Tests include generated lease/checkpoint histories, deterministic cancellation,
expiry and read-failure interleavings, and 60,000 indexed rows appended across
six file checkpoints. Independent-process Quack tests provide separate
concurrency coverage; generated sequential histories do not prove concurrent
correctness.

This compatibility path can be simplified only after the minimum supported
R/engine combinations pass direct native decoding tests, including nested NULLs,
empty collections, mixed lists, R class restoration and Quack round trips.
A development build passing the small reproducer is not sufficient to remove it.
The R descriptors remain necessary for distinctions such as data frames versus
lists even when the driver can decode every native value.
