---
title: "Serialized R objects in BLOBs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Serialized R objects in BLOBs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Native values remain inspectable in SQL. For objects outside the native mapping,
callers can serialize explicitly: raw vectors are stored as BLOB values.
CanardAbsurd does not choose a codec, unpack bytes, or fall back to serialization
when native admission fails.

These examples use a single-process, in-memory database. Shared deployments use
`ca_serve()` and `ca_connect()`; the same raw-vector contract applies over Quack.
The integration tests exercise serialized checkpoints in independent R workers.

## Base R serialization

This object contains a matrix and a language object, which are not admitted
as native workflow values. The handler uses a known format and returns serialized
bytes from its checkpoint.

```{r rds-workflow}
library(CanardAbsurd)
db <- ca_open()
object <- list(matrix = matrix(1:6, 2L), expression = quote(x + 1L))
id <- ca_spawn(db, "rds-v3", serialize(object, NULL, version = 3L))
outcome <- ca_run(ca_claim(db), function(input, task) {
  ca_step(task, "transform", function() {
    value <- unserialize(input)
    value$matrix <- value$matrix * 2L
    serialize(value, NULL, version = 3L)
  })
})
unserialize(outcome$result)
```

The database can inspect the outer format and byte count, not the R object's
contents. Whole-value materialization avoids the indexed VARIANT extraction
issue described in [Native values and DuckDB R compatibility](native-values.html).

```{r blob-inspection}
DBI::dbGetQuery(db@con, "
  WITH payload AS MATERIALIZED (
    SELECT input FROM canard_absurd.tasks WHERE id = ?
  )
  SELECT variant_typeof(input) AS native_type,
         octet_length(input::BLOB) AS bytes
  FROM payload
", params = list(id))
```

## qs2

`qs2` also produces a raw vector. The package remains unaware of the encoding;
the worker selects the matching decoder. Thread use is explicit here.

```{r qs2-workflow}
packed <- qs2::qs_serialize(object, nthreads = 1L)
ca_spawn(db, "qs2-v1", packed)
outcome <- ca_run(ca_claim(db), function(input, task) {
  ca_step(task, "copy", function() {
    value <- qs2::qs_deserialize(input, nthreads = 1L)
    qs2::qs_serialize(value, nthreads = 1L)
  })
})
qs2::qs_deserialize(outcome$result, nthreads = 1L)
```

## Sakura and reference-object reconstruction

[Sakura](https://github.com/RConsortium/sakura) augments R's reference hooks.
A class-specific serializer produces bytes, and a matching constructor rebuilds
the reference object in the receiving process. This is useful for objects such
as Arrow tables and torch tensors, whose external pointers cannot simply be
copied into another R session.

Here, an environment owns a buffered connection. Its hook records the source
bytes and cursor position, rather than the connection handle. Reconstruction
opens a fresh connection and seeks to that position.

```{r sakura-configuration}
cfg <- sakura::serial_config("buffer_reader",
  function(reader) {
    serialize(list(bytes = reader$bytes, position = seek(reader$con)), NULL)
  },
  function(bytes) {
    saved <- unserialize(bytes)
    reader <- new.env(parent = emptyenv())
    reader$bytes <- saved$bytes
    reader$con <- rawConnection(saved$bytes, "rb")
    seek(reader$con, saved$position)
    class(reader) <- "buffer_reader"
    reader
  })

reader <- new.env(parent = emptyenv())
reader$bytes <- charToRaw("alpha\nbeta\n")
reader$con <- rawConnection(reader$bytes, "rb")
class(reader) <- "buffer_reader"
readLines(reader$con, n = 1L)
packed <- sakura::serialize(reader, cfg)
close(reader$con)
```

The handler explicitly owns reconstruction and cleanup. The saved step is a
native string even though the input was serialized.

```{r sakura-workflow}
ca_spawn(db, "buffer-reader-v1", packed)
outcome <- ca_run(ca_claim(db), function(input, task) {
  ca_step(task, "line", function() {
    restored <- sakura::unserialize(input, cfg)
    on.exit(close(restored$con))
    readLines(restored$con, n = 1L)
  })
})
outcome$result
ca_close(db)
```

## What the application must retain

- Keep the serializer format/version and the hook-configuration identity with
  the job or its versioned handler. Sakura's stream selects hooks by position;
  reordering a configuration can change which decoder receives existing bytes.
- Receiving workers need compatible R packages and decoders. Sakura uses native
  binary serialization; test the architectures you intend to mix.
- A successful serialization call does not prove that a reference object is
  usable after restoration. Test its operations in a fresh process. Hooks must
  define what is reconstructed, including resource state and sharing semantics.
- Reconstructing a resource does not resume a running transaction, socket session
  or call stack. External files, services and artifacts referenced by a recipe
  must still exist.
- Only deserialize trusted payloads. R objects and reconstruction hooks can
  carry executable behavior. An API accepting arbitrary bytes must not turn
  them into arbitrary worker code through an unrestricted decoder.
- Reuse the original bytes for idempotent resubmission. Equivalent R objects are
  not guaranteed to serialize identically across versions or configurations.

Writes currently pass SQL text through Quack. BLOB literals use hexadecimal,
which represents each byte with two characters and also requires temporary R
allocations. Removing a byte quota does not make this an efficient bulk-data
channel. Large datasets can remain in an artifact store or dedicated tables,
with the job carrying references.

For a targets execution backend, this supplies a transport for execution objects
without forcing every R class into a SQL mapping. It does not itself implement
targets scheduling, cache validity, result acknowledgement or controller recovery.
