---
title: "A Quack server and R workers"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, cache = FALSE, collapse = TRUE,
                    comment = "#>", error = FALSE, message = FALSE)
```

Quack is required for a shared workflow database. One server process owns the
writable DuckDB file; producers and workers use their own
in-memory DuckDB connections to send SQL through Quack. They must not open the
server's database file directly: multi-process file access causes DuckDB/WAL
contention and locking errors. R handlers execute in those clients, never in a
server callback.

This article starts a temporary server and executes a worker over Quack. It
puts both handles in one R process so the evaluated guide does not need to start
a separate shell process; the connections still communicate through a Quack
endpoint. Production deployment separates the server and
workers as described below. This network-dependent article is built by pkgdown,
while the package vignettes run offline.

## Install the runtime

Install DuckDB's JSON and Quack extensions explicitly before serving or
connecting. From a source checkout, `make quack` performs this setup. In another
R environment, run this once with network access:

```r
dir.create(path.expand("~/.duckdb"), recursive = TRUE, showWarnings = FALSE)
local({
  con <- DBI::dbConnect(duckdb::duckdb())
  on.exit(DBI::dbDisconnect(con, shutdown = TRUE))
  DBI::dbExecute(con, "INSTALL json")
  DBI::dbExecute(con, "INSTALL quack")
})
```

The extensions are stored in DuckDB's shared extension directory so later R
processes using the same DuckDB runtime can load them. Package loading and
connections never download extensions.

## Run the server/client path

The demonstration uses a disposable localhost token. Use a securely supplied
nonempty token and a controlled network address for a deployment. Here `path`
is a temporary database file and `uri` is an available localhost endpoint.

```{r server}
library(CanardAbsurd)
path <- tempfile(fileext = ".duckdb")
withr::defer(unlink(c(path, paste0(path, ".wal"))))
uri <- sprintf("quack:127.0.0.1:%d", parallelly::freePort())
server <- ca_serve(path, uri, token = "article-local-only")
withr::defer(ca_close(server))
```

```{r client}
worker <- ca_connect(uri, token = "article-local-only")
withr::defer(ca_close(worker))
```

```{r server-worker}
sum_values <- function(input, ctx) {
  ca_step(ctx, "sum", function() sum(unlist(input$values)))
}
id <- ca_spawn(worker, "sum", list(values = list(10, 20, 12)),
               id = "remote-sum-001", queue = "reports")
delays <- c(0.01, 0.05, 0.1)
withCallingHandlers(
  ca_work(worker, list(sum = sum_values), queue = "reports", max_tasks = 1),
  canard_retryable = function(conflict) {
    if (conflict$attempts <= length(delays)) {
      Sys.sleep(delays[[conflict$attempts]])
      invokeRestart("canard_retry")
    }
  }
)
task <- ca_inspect(worker, id)
list(runtime = ca_runtime(worker), state = task$state, result = task$result)
```

These are independent database connections in one R process, communicating
through a Quack endpoint. The `tinytest` suite also executes workers in
independent R processes and kills worker and database-host processes to test
recovery.

## Deploy across processes

Use the same `ca_serve()` call in a long-lived server R process, with a persistent
database path. It is the only process that opens that file and it must remain
alive to serve requests. Every producer and worker calls `ca_connect()` with the
endpoint and token; workers register their handler lists and enter `ca_work()`.

Each worker executes one attempt at a time. `lease_seconds` must cover the time
between named checkpoint calls or explicit heartbeats. `poll_seconds` controls idle
polling, and `idle_timeout` controls when an idle worker returns. A worker that
only knows one handler will not claim other task names.

SQL retries are independent of the task's failure budget. The calling handler
above permits three retries per conflicting statement, with delays chosen by
the caller. The restart repeats only that SQL statement, not the R handler or
its external effects. Storage errors include the operation, total attempts,
elapsed seconds, and the original DBI condition. Transport failures propagate
even when raised by a step inside a handler and offer no retry restart.
Released drivers require a message-based compatibility adapter for conflict
classification. The condition's `message_based` field discloses that path; see
[error metadata compatibility](durability.html#error-metadata-compatibility)
for its limits and the upstream reports.

## Inspect and stop

`ca_inspect()` reports state, attempts, failures, available time, owner, and lease
deadline. `ca_runtime()` reports the executing server's versions. `ca_close()`
closes a client; closing the server handle stops its endpoint and closes the file.

```{r close}
ca_close(worker)
ca_close(server)
```

Keep handles process-local. Give package operations their dedicated connections
rather than wrapping them in externally managed transactions.

## Protect the endpoint

The token grants trusted database access, not per-queue tenant isolation. Use
network restrictions and TLS at authenticated ingress for non-local deployments.
Do not expose an unprotected Quack server to the internet. Package loading and
connections never install extensions; extension acquisition is explicit setup.
