---
title: "A Quack server and R workers"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, cache = FALSE, collapse = TRUE,
                    comment = "#>", error = FALSE, message = FALSE)
library(CanardAbsurd)
path <- tempfile(fileext = ".duckdb")
withr::defer(unlink(c(path, paste0(path, ".wal"))))
uri <- sprintf("quack:127.0.0.1:%d", parallelly::freePort())
```

One process owns the writable DuckDB file. Quack exposes remote SQL access;
clients use their own in-memory DuckDB connection to reach it. R handlers execute
in those clients, not in server callbacks.

This article starts a real temporary server and executes a worker over Quack.
All R chunks run during the site build. Quack must be explicitly installed for
the R DuckDB runtime before rendering; `make quack` provides that setup in this
checkout. This network-dependent article is built by pkgdown, while the package
vignettes run offline.

## Run the server/client path

The demonstration uses a disposable localhost token. Use a securely supplied
nonempty token and a controlled network address for a deployment.

Here `path` is a temporary database file and `uri` is an available localhost
endpoint chosen for the render.

```{r server}
server <- ca_serve(path, uri, token = "article-local-only")
```

```{r server-cleanup, include=FALSE}
withr::defer(ca_close(server))
```

```{r client}
worker <- ca_connect(uri, token = "article-local-only")
```

```{r client-cleanup, include=FALSE}
withr::defer(ca_close(worker))
```

```{r server-worker}
sum_values <- function(input, ctx) {
  ca_step(ctx, "sum", function() sum(unlist(input$values)))
}
id <- ca_spawn(worker, "sum", list(values = list(10, 20, 12)),
               id = "remote-sum-001", queue = "reports")
delays <- runif(3, 0, 0.1)
withCallingHandlers(
  ca_work(worker, list(sum = sum_values), queue = "reports", max_tasks = 1),
  canard_retryable = function(conflict) {
    if (conflict$attempts <= length(delays)) {
      Sys.sleep(delays[[conflict$attempts]])
      invokeRestart("canard_retry")
    }
  }
)
task <- ca_inspect(worker, id)
list(runtime = ca_runtime(worker), state = task$state, result = task$result)
```

These are independent database connections in one R process, communicating
through a real Quack endpoint. The `tinytest` suite also executes workers in
independent R processes and kills worker and database-host processes to test
recovery.

## Deploy across processes

Use the same `ca_serve()` call in a long-lived server R process, with a persistent
database path. The process must remain alive to serve requests. Worker processes
call `ca_connect()` with that endpoint and token, register their handler lists,
and enter `ca_work()`.

Each worker executes one attempt at a time. `lease_seconds` must cover the time
between step boundaries or explicit heartbeats. `poll_seconds` controls idle
polling, and `idle_timeout` controls when an idle worker returns. A worker that
only knows one handler will not claim other task names.

SQL retries are independent of the task's failure budget. The calling handler
above permits three retries per conflicting statement, with delays chosen by
the caller. The restart repeats only that SQL statement, not the R handler or
its external effects. Storage errors include the operation, total attempts,
elapsed seconds, and the original DBI condition. Transport failures propagate
even when raised by a step inside a handler and offer no retry restart.
Released drivers require a message-based compatibility adapter for conflict
classification. The condition's `message_based` field discloses that path; see
[error metadata compatibility](durability.html#error-metadata-compatibility)
for its limits and the upstream reports.

## Inspect and stop

`ca_inspect()` reports state, attempts, failures, available time, owner, and lease
deadline. `ca_runtime()` reports the executing server's versions. `ca_close()`
closes a client; closing the server handle stops its endpoint and closes the file.

```{r close}
ca_close(worker)
ca_close(server)
```

Keep handles process-local. Give package operations their dedicated connections
rather than wrapping them in externally managed transactions.

## Protect the endpoint

The token grants trusted database access, not per-queue tenant isolation. Use
network restrictions and TLS at authenticated ingress for non-local deployments.
Do not expose an unprotected Quack server to the internet. Package loading and
connections never install extensions; extension acquisition is explicit setup.
