---
title: "A Quack server and R workers"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, cache = FALSE, collapse = TRUE, comment = "#>", error = FALSE)
library(CanardAbsurd)
```

One process owns the writable DuckDB file. Quack exposes remote SQL access;
clients use their own in-memory DuckDB connection to reach it. R handlers execute
in those clients, not in server callbacks.

This article starts a real temporary server and executes a worker over Quack.
All R chunks run during the site build. Quack must be explicitly installed for
the R DuckDB runtime before rendering; `make quack` provides that setup in this
checkout. This network-dependent article is built by pkgdown, while the package
vignettes run offline.

## Run the server/client path

The demonstration uses a disposable localhost token. Use a securely supplied
nonempty token and a controlled network address for a deployment.

```{r server-worker}
local({
  path <- tempfile(fileext = ".duckdb")
  on.exit(unlink(c(path, paste0(path, ".wal"))), add = TRUE)
  uri <- sprintf("quack:127.0.0.1:%d", parallelly::freePort())

  server <- ca_serve(path, uri, token = "article-local-only")
  on.exit(ca_close(server), add = TRUE, after = FALSE)
  producer <- ca_connect(uri, token = "article-local-only")
  on.exit(ca_close(producer), add = TRUE, after = FALSE)
  worker <- ca_connect(uri, token = "article-local-only")
  on.exit(ca_close(worker), add = TRUE, after = FALSE)

  id <- ca_spawn(producer, "sum", list(values = list(10, 20, 12)),
                 id = "remote-sum-001", queue = "reports")
  handlers <- list(sum = function(input, ctx) {
    ca_step(ctx, "sum", function() sum(unlist(input$values)))
  })
  ca_work(worker, handlers, queue = "reports", max_tasks = 1)

  task <- ca_inspect(producer, id)
  stopifnot(task$state == "completed", task$result == 42)
  list(runtime = ca_runtime(worker), state = task$state, result = task$result)
})
```

These are independent database connections in one R process, communicating
through a real Quack endpoint. The `tinytest` suite also executes workers in
independent R processes and kills worker and database-host processes to test
recovery.

## Deploy across processes

Use the same `ca_serve()` call in a long-lived server R process, with a persistent
database path. The process must remain alive to serve requests. Worker processes
call `ca_connect()` with that endpoint and token, register their handler lists,
and enter `ca_work()`.

Each worker executes one attempt at a time. `lease_seconds` must cover the time
between step boundaries or explicit heartbeats. `poll_seconds` controls idle
polling, and `idle_timeout` controls when an idle worker returns. A worker that
only knows one handler will not claim other task names.

## Inspect and stop

`ca_inspect()` reports state, attempts, failures, available time, owner, and lease
deadline. `ca_runtime()` reports the executing server's versions. `ca_close()`
closes a client; closing the server handle stops its endpoint and closes the file.

Keep handles process-local. Give package operations their dedicated connections
rather than wrapping them in externally managed transactions.

## Protect the endpoint

The token grants trusted database access, not per-queue tenant isolation. Use
network restrictions and TLS at authenticated ingress for non-local deployments.
Do not expose an unprotected Quack server to the internet. Package loading and
connections never install extensions; extension acquisition is explicit setup.
